---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dpi = 600,
  dev = "png",
  out.width = "100%"
)
```

# pbcusol

<!-- badges: start -->
<!-- badges: end -->

`pbcusol` predicts equilibrium lead and copper solubility using the US EPA databases LEADSOL (Schock et al. 1996) and CU2SOL (Schock et al., 1995), along with those available as a part of PHREEQC (Charlton and Parkhurst, 2011; Parkhurst and Appelo, 2013). `pbcusol` uses `tidyphreeqc` (Dunnington, 2019)---a convenient interface for PHREEQC in R---for most solubility computations.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("bentrueman/pbcusol")
```
## Example

For this example, you will need the `tidyverse` family of packages, along with `plyr::round_any()`, and `viridis::scale_fill_viridis()`.

```{r load, message=FALSE}
library("pbcusol")
library("tidyverse")
library("viridis")
```

```{r theme, echo=FALSE}
theme_set(theme_classic() + theme(legend.position = "right", strip.background = element_blank()))
```

Use `cu_sol()` to calculate the equlibrium solubility of multiple copper phases that occur in drinking water systems, over a wide range of pH values and dissolved inorganic carbon concentrations. N.B., evaluate on a smaller grid to speed this up!

```{r cu-sol}
solutions_cu <- list("Cu(OH)2", "Tenorite", "Malachite") %>%
  set_names() %>%
  map_dfr(
    ~cu_sol(
      ph = seq(6, 11, by = .025),
      dic = seq(1, 150, by = .5),
      phase = .x
    ),
    .id = "phase"
  )
```

Plot the data (n.b., plots shown here will differ slightly from those the following code generates).

```{r cu-plot-display, eval=FALSE}
solutions_cu %>% 
  mutate(
    log10_cu_ppb = log10(cu_ppb),
    dic_ppm = plyr::round_any(dic_ppm, .5)
  ) %>% 
  ggplot(aes(x = dic_ppm, y = pH)) + 
  facet_wrap(vars(phase)) + 
  geom_raster(aes(fill = log10_cu_ppb)) +
  geom_contour(aes(z = log10_cu_ppb), col = "white") +
  viridis::scale_fill_viridis(option = "magma")
```

```{r cu-plot, fig.height=3, echo=FALSE}
solutions_cu %>% 
  mutate(
    phase = fct_recode(phase, "Cu(OH)[2]" = "Cu(OH)2"),
    cu_ppb = log10(cu_ppb),
    dic_ppm = plyr::round_any(dic_ppm, .5)
  ) %>% 
  ggplot(aes(x = dic_ppm, y = pH)) + 
  facet_wrap(vars(phase), labeller = label_parsed) + 
  geom_raster(aes(fill = cu_ppb)) +
  geom_contour(aes(z = cu_ppb), col = "white", linetype = 2, size = .2, breaks = seq(-4, 5, by = .1)) + 
  geom_contour(aes(z = cu_ppb), col = "white", breaks = -3:5) + 
  viridis::scale_fill_viridis(option = "magma", labels = function(breaks) 1e-3 * (10 ^ breaks)) +
  labs(
    x = expression("[C]"[inorganic]~"(mg L"^"-1"*")"),
    y = "pH",
    fill = expression("[Cu] (mg L"^"-1"*")")
  )
```

Use `pb_sol()` to do the same for lead.

```{r pb-sol}
solutions_pb <- tibble(
  phase = c("Cerussite", "Hydcerussite", "Hxypyromorphite"),
  phosphate = c(0, 0, 0.33)
) %>%
  mutate(
    data = map2(
      phase, phosphate,
      ~ pb_sol(
          ph = seq(6, 11, by = .025),
          dic = seq(1, 50, by = .5),
          phosphate = .y,
          phase = .x
      )
    )
  ) %>% 
  unnest(data)
```

Plot the data:

```{r pb-plot-display, eval=FALSE}
solutions_pb %>% 
  mutate(
    log10_pb_ppb = log10(pb_ppb),
    dic_ppm = plyr::round_any(dic_ppm, .5)
  ) %>% 
  ggplot(aes(x = dic_ppm, y = pH)) + 
  facet_wrap(vars(phase)) + 
  geom_raster(aes(fill = log10_pb_ppb)) +
  geom_contour(aes(z = log10_pb_ppb), col = "white") +
  viridis::scale_fill_viridis(option = "magma")
```

```{r pb-plot, fig.height=3, echo=FALSE}
solutions_pb %>% 
  mutate(
    phase = fct_recode(phase, "Hydroxylpyromorphite" = "Hxypyromorphite", "Hydrocerussite" = "Hydcerussite"),
    pb_ppb = log10(pb_ppb),
    dic_ppm = plyr::round_any(dic_ppm, .5)
  ) %>% 
  ggplot(aes(x = dic_ppm, y = pH)) + 
  facet_wrap(vars(phase)) + 
  geom_raster(aes(fill = pb_ppb)) +
  geom_contour(aes(z = pb_ppb), col = "white", linetype = 2, size = .2, breaks = seq(-4, 5, by = .1)) + 
  geom_contour(aes(z = pb_ppb), col = "white", breaks = -3:5) + 
  viridis::scale_fill_viridis(option = "magma", labels = function(breaks) 1e-3 * (10 ^ breaks)) +
  labs(
    x = expression("[C]"[inorganic]~"(mg L"^"-1"*")"),
    y = "pH",
    fill = expression("[Pb] (mg L"^"-1"*")")
  )
```

The helper function `calculate_dic()` may be useful for converting alkalinity to dissolved inorganic carbon.

# References 

Charlton, S.R., and Parkhurst, D.L, 2011, Modules based on the geochemical model PHREEQC for use in scripting and programming languages: Computers & Geosciences, v. 37, p. 1653-1663.

Dunnington, D. 2019. tidyphreeqc: Tidy Geochemical Modeling Using PHREEQC. https://github.com/paleolimbot/tidyphreeqc.

Parkhurst, D. L., and C. A. J. Appelo. 2013.  Description of input and examples for PHREEQC version 3--A computer program
for speciation, batch- reaction, one-dimensional transport, and inverse geochemical calculations: U.S. Geological
Survey Techniques and Methods, book 6, chap. A43, 497 p. http://pubs.usgs.gov/tm/06/a43.

Schock, M. R., D. A Lytle, and J. A. Clement. 1995. “Effect of pH, DIC, orthophosphate and sulfate on drinking water cuprosolvency.” National Risk Management Research Lab., Cincinnati, OH (United States).

Schock, M. R., I. Wagner, and R. J. Oliphant. 1996. “Corrosion and solubility of lead in drinking water.” In Internal corrosion of water distribution systems, 2nd ed., 131–230. Denver, CO: American Water Works Association Research Foundation.

